Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals

### 2022-05-19 draft_BLC/BLC_AM-ID_manuscript.tex
manuscript revision: major edits, figures

\usepackage{graphicx} %for images
\graphicspath{{images/}}
\usepackage{caption}
\captionsetup[table]{skip=0pt,singlelinecheck=off}
\captionsetup[figure]{skip=0pt,singlelinecheck=off}

@@

\newpage
\begin{myparindent}{0pt}
Abstract:\\
This study investigates the phonetic outcome of mixed-language speech in Bengali and Indian English, towards understanding cross-language interaction in highly proficient bilinguals.  We compare spectral properties of L2 vowels \nt{\ae} (common to L1, L2) and \nt{2} (absent in L1) in unilingual vs. mixed productions. Results reveal dynamic and asymmetrical changes in both vowels;  \nt{2} shifting towards a related L1 category \nt{a:}. We interpret this as a temporary increase in cross-language phonetic interaction during mixed-language use, leading to a shift towards L1 norms, evidencing transfer effects on L2 vowels for the first time. We elicit mixed productions through two tasks: cued picture-naming and code-switching, to assess if experimental paradigm independently influences the behavior under study. Results reveal no difference across paradigms. We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction, postulated discursive factors in code-switching, and the issue of comparability between paradigms in transfer studies.\\

@@

\section{Introduction: Phonetic transfer}

@@

The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India. We measure spectral properties: F1 and F2, of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2. Mixed-language data is elicited in two switching paradigms: cued picture-naming and code-switching. We compare these results to assess if differences between the paradigms independently influence the outcome of phonetic interaction. The results demonstrate a dynamic effect of transfer on L2 vowels during mixed-language production, \alert{but no difference in the extent of transfer between the two paradigms --and slightly different patterns of shift among the paradigms?}. We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.

@@

\alert{note that there are large regional variations in Indian English, and the literature on IEphonology is thus not "standardized" or uncontested. For the present study, we focus on a variety of IE spoken by an educated elite, used in news channels, taught in classrooms (citations)}

Recent research comparing the segmental and suprasegmental properties of IE spoken in different parts of the country report many commonalities (see \cite{sirsa2013effects} for a review). This suggests that in spite of the varied L1s of its speakers, IE has a target phonology that is distinct from any of these, as well as from other native varieties of English. Thus, it is fruitful to think of regional variations as resulting from L1-influence on a common underlying target. Excepting a small minority of specific communities \citep{pandey201517, wells1982accents, coelho1997anglo}, there are very few L1 speakers of IE.

@@

\subsubsection*{Vowel systems} \label{vowel systems}
The vowel inventory of Western Bangla (the variety spoken by the participants in this study) consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \citep{garry2001facts}. Note that there is no mid-central vowel category.

The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2, a:, O, o, U, u}, represented by the lexical set KIT, FLEECE, DRESS, FACE, and TRAP, STRUT, PALM, LOT/CLOTH, GOAT, FOOT, GOOSE \citep{wells1982accents, masica1972sound}. %Their distribution in the F1XF2 space is shown in (insert table 1.7, Pandey). 
A single mid-central vowel corresponds to the categories \nt{2,@,3:}, which are treated as distinct in many native varieties of English \citep{nihalani1979indian,wells1982accents,hickey2005legacies,bansal1969intelligibility}. Since the English items used in this study are traditionally transcribed with \nt{2}, we use this symbol to indicate the mid-central vowel throughout.

@@

\alert{discussion: more about ICM, assumptions about processing}

@@

In the present study, our research question concerns whether L2 vowels are produced differently in unilingual vs mixed conditions. We do not elicit unilingual Bengali productions. To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we used production data from the freely available Bengali corpus, SHRUTI \citep{shruticorpus}. Formant values were extracted from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, to get formant estimates from the steady state, which should most closely approximate the production target. Next, we compared these to our participants' productions in the unilingual English condition to understand the relative positions of the categories. \alert{fig. .. plots-- from current plot- remove the dotted line. Filter by context, then remove group=context from figure code} F1-F2 plots show that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, than the English categories \nt{\ae} and \nt{2} respectively.

@@

\subsection{Participants} \label{participants}

10 Bengali-English bilingual speakers (male=5, age range 19 to 28), completed the tasks. All the participants speak other languages in addition to Bengali and English. At the time of recording, they were living in or around the EFL University campus in Hyderabad, which has a linguistically diverse population. Speakers were selected on the basis of responses to a Language Background Questionnaire to ensure comparable LSRW (listening, speaking, reading, writing) skills in L1 and L2 across participants, and were compensated for their time.

\subsection{Stimuli}
Twenty monosyllabic English words containing the vowels \nt{2} or \nt{\ae} were selected as target words. To minimize coarticulatory effects on the vowel, the pre-vocalic consonants were all voiced plosives (\nt{b} or \nt{d}). 10 monosyllabic English words that do not contain the target vowel served as filler items. We avoided English words that are lexicalized loanwords in Bengali. %The list of words can be found in sec.\ref{appendix}.

\subsection{Procedure}
Participants were recorded individually in a sound attenuated room. %, using a ... microphone type/description?
Productions were recorded in Praat \citep{boersma2016praat} at a 44.1K sampling rate. Every participant performed four iterations each of two tasks-- cued picture-naming and code-switching. Utterances in the two conditions (unilingual and mixed) were randomly interspersed within each iteration. In the unilingual condition, the target word was preceded by another English word in the utterance. In the mixed condition, the target was preceded by a Bengali word and thus produced as a switch from Bengali. Every target word appeared twice during a test block; once in a unilingual and once in a mixed condition. Participants alternated between the two tasks, and were allowed to take a short break after each test block. The tasks are described below:

@@

The sound preceding the target word was uniform across all sentences (the final sound in \textit{yellow} is produced as [o] in Indian English), to minimize any possible differences due to coarticulatory effects on the target vowel. A specifier-noun construction was used to ensure identical word order in both languages, giving a uniform prosodic context for the target word across condition. In presenting the mixed-language utterances, we used mixed orthography. \\

@@

All statistical analyses were performed in R \citep{r}. We used linear mixed effects models to test the effect of experimental conditions on vowel height (F1) and backness (F2) respectively. Since formants were extracted from five points over the course of articulating the vowel, time was coded numerically (1-5), and treated as an independent variable. This was scaled and centered prior to analysis and treated as a continuous predictor. We fitted models using the \emph{lme4} package \citep{lme4}, and used \emph{lmerTest} \citep{lmerTest} to obtain estimates of p values for individual variables. For all mixed models, the alpha criterion was set at $|t| > 2 $. The dependent variables were normalized F1 and F2, and we fitted separate models for these. The independent variables or fixed factors, along with their levels, are given in table \ref{table variables} \alert{add variable type with each: factor, numerical, ...}. Subject and Item were treated as random factors.\\

@@

Our study involved two IE vowels. Vowel category is expected to systematically affect formant frequencies (FFs).

The experimental treatment of interest is Context, i.e. whether the target item was presented in a unilingual (English) or mixed (Bengali) context. Our main research question (\bf{L2 vowels}) asks whether L2 vowel quality (formant frequencies) of proficient bilinguals is affected by the Context of utterance. We hypothesize that the direction of shift in the mixed condition is towards the corresponding L1 category. Thus, we expect the effect of Context on formant frequencies, i.e. the direction of shift, to be affected by the vowel category (c.f. section \ref{asymmetry between sounds}). Thus, our main effect of interest is the interaction term Context*Vowel. This corresponds to the transfer pattern.

Our third research question (\bf{Paradigm}) asks whether the language-mixing paradigm, or Task, affects the extent and direction of transfer. If this is the case, we expect Task to interact with the transfer term, giving a three way Task*Context*Vowel interaction.

Since we measured FFs at five points in the vowel, Time is expected to affect the FF. However, the precise effect of formant dynamics is expected to depend on the vowel category. Thus, we expect the interaction of Vowel*Time to predict formant frequency. If transfer patterns change across time, then we should expect it to interact with the transfer term, giving a Context*Vowel*Time interaction.

@@

First, to test hypothesis 1 (\bf{L2 transfer}), we fitted a model with vowel*time, and vowel*context as fixed effects. The transfer term (vowel*context) is significant in this model (model coefficient/ unstandardized effect size ($\beta$)= -2.330e-01, standard error (SE)=  5.244e-02, t= -4.44, p=0.0002). Comparing this to a null model that lacks the transfer term shows that the full model is significantly better ($\chi^2(2) = 14.42, p=  0.0007$). To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction. The full model was better ($\chi^2(1) = 12.51, p= 0.0004$).

To test the hypothesis \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects. This three-way interaction term is significant ($\beta$= -1.955e-01, SE=  5.453e-02, t= -3.59, p= 0.0003). We also find a significant effect of task on FFs ($\beta$=3.300e-01, SE=2.668e-02, t=12.371, p=2e-16). Comparing this model to null models that lack the fixed effect Task ($\chi^2(4) = 572.61, p=  2.2e-16$) shows that the full model is better. Because Task has an independent effect on F1, and adding interactions increases model parameters, we fitted a second null model with Task as a separate fixed effect without the interaction term to test whether the interaction term is needed. However, the full model was still a better fit for the data ($\chi^2(3) = 23.82, p=  2.722e-05$). Next, we tested two other null models to verify that the full model with the three-way interaction term is indeed the optimal model for the data, by replacing the three-way interaction term with: (i) vowel*context + task*context (to test if task only interacts with context, rather than the transfer term vowel*context). The full model is better ($\chi^2(2) = 23.28, p= 8.78e-06$); (ii) vowel*context + task*vowel (to test if task only interacts with vowel, rather than the transfer term vowel*context). The full model is better ($\chi^2(2) = 13.36, p= 0.001$). This shows that the three-way interaction is the best fit for the data. Because an independent effect of task on FFs is unexpected (c.f. ), we wanted to see if the effect might be local, i.e. decrease over time. We added a task*time interaction term to the full model, which is significant ($\beta$= -1.216e-01, SE=  1.347e-02, t= -9.028, p= 2e-16 ) and improves model fit ($\chi^2(1) = 81.22, p= 2.2e-16$).

@@

Thus, the optimal model was: f1 ~ time*vowel + vowel*context*task + time*task + (1 + context|word) + (1 + context + vowel|subject). Table \ref{table_f1} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model. \alert{p-vales smaller than 0.05 are in bold.}

\subsubsection*{Interpretation}
We extracted formant estimates from five uniformly spaced points in the vowel, from the time at 5\% into the vowel (at the start of articulation) to 45\% (at the mid-part, or steady state, which is expected to correspond to the articulatory target for the category). As expected, we see that F1 values are significantly predicted by Time in the model, reflecting th dynamic nature of articulation which causes acoustic variation over the course of time. The positive slope of the estimate for Time indicates that F1 increases, i.e. the two categories in the study show lowering over the course of articulation. Moreover, the significant Time*Vowel interaction shows that the F1 trajectory is different for the two vowels. This is expected, as the categories are expected to present different articulatory targets.

 Vowel is not a significant predictor of F1. This appears surprising, given that the phonological category is expected to systematically affect vowel quality. However, a boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two categories \nt{2, \ae} and the extent of height difference between them. Note that these measures are only from the English (unilingual) condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching. Given that estimates for fixed effects are pooled across participants, this explains the lack of reliable differences between \nt{2} and \nt{\ae}. The negative slope of the estimate for Context shows that F1 is lower, i.e. vowels are higher, in the unilingual (e) condition compared to the mixed (b) condition. Put more intuitively, vowels are lowered in the mixed condition. However, the lack significance indicates that using English words in mixed utterances does not lead to a uniform lowering/raising across vowels. Instead, the significant interaction between Vowel and Context signals an asymmetrical shift in the categories \nt{2} and \nt{\ae} in the mixed-language condition. Specifically, the Vowel*Context interaction has a negative slope, showing that the lowering is greater in \nt{2}.

The results also reveal a significant effect of Task, such that the sentence task (code-switching) leads to higher F1 values in both vowels relative to the cued picture-naming task as: vowels are lowered in the code-switching task, independent of experimental condition. This is surprising, as we have no aproiri reason to expect a uniform difference in vowel quality across tasks. A significant interaction with Time with a positive slope of the estimate shows that this effect is local: it is strongest at the beginning of the vowel, and reduces over the course of the vowel. This is consistent with our assumption that the unexpected Task effect is an artifact of the paradigm, possibly resulting from coarticulatory effects with the preceding sound in the carrier phrase. Of greater interest to us is the significant Vowel*Context*Task interaction, which indicates a difference in transfer patterns across paradigms. Specifically, the negative slope of the estimate suggests that transfer in the F1 (height) dimension are greater in the code-switching (s) task compared to the picture-naming (p) task.\\

@@

Because an independent effect of task on FFs is unexpected \alert{(c.f. next section)}, we wanted to see if the effect might be local, i.e. decrease over time. We added a task*time interaction term to the full model, which is significant ($\beta$= 6.237e-02, SE=  8.876e-03, t= 7.027, p= 2.23e-12 ) and improves model fit ($\chi^2(1) = 49.271, p= 2.23e-12$).

@@

f1: contexte had -ve slope, vt also has -ve slope, so vt is moving up in e ie down in b condition. vowel*context has -ve slope, which means that vt is moving more than ae. Now, vowel*context*task has -ve slope, which means in the s task, compared to p, there is  more downward movement. So more transfer in s task.

>>>>

\usepackage{graphicx} %for images
\graphicspath{{images/}}
\usepackage{caption}
\captionsetup[table]{skip=0pt,singlelinecheck=off}
\captionsetup[figure]{skip=0pt,singlelinecheck=off}
\usepackage{subcaption}

@@

\newpage
\begin{myparindent}{0pt}
Abstract:\\
This study investigates the phonetic outcome of mixed-language speech in Bengali and Indian English, towards understanding cross-language interaction in highly proficient bilinguals.  We compare spectral properties of L2 vowels \nt{\ae} (common to L1, L2) and \nt{2} (absent in L1) in unilingual vs. mixed productions. Results reveal dynamic and asymmetrical changes in both vowels;  \nt{2} shifting towards a related L1 category \nt{a:}. We interpret this as a temporary increase in cross-language phonetic interaction during mixed-language use, leading to a shift towards L1 norms, evidencing transfer effects on L2 vowels for the first time. We elicit mixed productions through two tasks: cued picture-naming and code-switching, to assess if experimental paradigm independently influences the behavior under study. Results reveal parallel patterns, but a slight difference in magnitude of transfer across paradigms. We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction, postulated discursive factors in code-switching, and the issue of comparability between paradigms in transfer studies.\\

@@

\section{Introduction: Phonetic transfer}\label{introduction}

@@

The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India. We measure spectral properties: F1 and F2, of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2. Mixed-language data is elicited in two switching paradigms: cued picture-naming and code-switching. We compare these results to assess if differences between the paradigms independently influence the outcome of phonetic interaction. The results demonstrate a time-invariant effect of transfer on L2 vowels during mixed-language production, and parallel pattern but different magnitude of transfer between the two paradigms. We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.

@@

Note that ``Indian English" is not a monolithic entity; there are large regional variations in the English speech of Indian speakers, and the literature on IE phonology is thus neither ``standardized" nor uncontested. For the present study, we focus on a variety of IE spoken by an educated elite, used in news channels, often taught as a target in classrooms, and generally considered to be free of any particular regional affiliation \citep[c.f.][who reports this variety to be ``a \textit{de-facto} norm", and thus calls it General(ized) Indian English]{masica1972sound}. This sound system is described in section \ref{vowel systems}. However, recent research comparing the segmental and suprasegmental properties of IE spoken in different parts of the country also report many commonalities (see \cite{sirsa2013effects} for a review). It has thus been suggested that in spite of the varied L1s of its speakers, IE has a target phonology that is distinct from any of these, as well as from other native varieties of English. Thus, it is fruitful to think of regional variations as resulting from L1-influence on a common underlying target. Excepting a small minority of specific communities \citep{pandey201517, wells1982accents, coelho1997anglo}, there are very few L1 speakers of IE. Thus, English is largely learned and used as an L2.

@@

\subsubsection*{Vowel systems} \label{vowel systems}
The vowel inventory of Western Bangla (the variety spoken by the participants in this study) consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \citep{garry2001facts}. The vowel chart in figure \ref{vowels_bengali} shows their distribution in the F1XF2 space. Note that there is no mid-central vowel category.

The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2/3:, a:, O, o, U, u}, represented by the lexical set KIT, FLEECE, DRESS, FACE, and TRAP, STRUT, PALM, LOT/CLOTH, GOAT, FOOT, GOOSE \citep{wells1982accents, masica1972sound}. The vowel chart in figure \ref{vowels_ie} shows their distribution in the F1XF2 space. 
A single mid-central vowel corresponds to the categories \nt{2,@,3:}, which are treated as distinct in many native varieties of English \citep{nihalani1979indian,wells1982accents,hickey2005legacies,bansal1969intelligibility}. Since the English items used in this study are traditionally transcribed with \nt{2}, we use this symbol to indicate the mid-central vowel throughout.

@@

@@

In the present study, our research question concerns whether L2 vowels are produced differently in unilingual vs mixed conditions. Previous literature on phonetic transfer suggests that such shifts are not random, but rather directional and targeted. Although our primary research question merely concerns whether productions in unilingual vs mixed conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories. To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we processed and used existing speech data from a freely available Bengali corpus, SHRUTI \citep{shruticorpus}. The procedure is described below:

\subsubsection{Baseline Bengali data from the SHRUTI corpus}

The SHRUTI corpus is the largest free corpus of Western Bengali speech that we are aware of. It comprises of 21.64 hours of read speech, containing 7383 unique sentences (22012 unique words) from Bengali newspapers in different domains, covering the most frequently used words of the language. The speakers are 34 adults (age range 20-40, 75\% male, 25\% female) who have grown up in the Indian state of West Bengal, and speak standard colloquial Bengali. All of the speakers have university-level education or higher. Given these parallels in linguistic, regional, and socio-economic background, we believe that Bengali speech from this corpus is representative of the average target Bengali vowel categories that are of interest in this study.
To estimate average vowel positions, we extracted formant values from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, to get estimates from the steady state, which should most closely approximate the production target. Next, we compared these to our participants' productions in the unilingual English condition to understand the relative positions of these categories. This is visualized in figure \ref{vowels_unilingual_ie_bengali}.

Note that we refrained from using corpus data for average Indian English vowel positions, opting instead to use the participant data from this study, because IE has high regional variability (c.f. discussion in section \ref{bengali_english_in_india}). To the best of our knowledge, there is no freely available corpus of IE speech from Bengali speakers. 
The F1XF2 plot indicates that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, than the English categories \nt{\ae} and \nt{2} respectively.

@@

\subsection{Participants} \label{participants}
Participants were recruited through an open call at the EFL University campus in Hyderabad, India, circulated through social media. Interested volunteers filled a brief language background questionnaire. These responses were used to screen participants, in the manner described below. 
As noted in \ref{bengali_english_in_india}, India has a super multilingual an linguistically diverse population, and EFLU being a central university represents an extremely diverse population, as students are from different parts of the country, having both different L1s, and different varieties of English. Our aim in screening participants was to minimize between-participant random variability, while maintaining ecological validity considering the actual day-to-day settings in which such behaviors likely take place. We chose two metrics to do this: (i) where the participant grew up; (ii) number of years of formal education in Bengali. The first ensures that the variety of Bengali that participants are exposed to in their surroundings is similar. This would mean that the target representations for Bengali vowels would be expected to be similar. This was especially important because we were not eliciting unilingual L1 data in this study, and were using a corpus to estimate the position of Benali vowels in the formant space. Based on previous literature (c.f. section \ref{asymmetry between sounds}), we expect that L2 vowels shift towards related L1 categories during mixed language utterances. Thus, we wanted to make sure that differences between participants in the pattern of shift were actually due to individual differences, rather than different acoustic targets. Another reason to control for place of growing up was that it is a fairly reliable predictor of what languages the person was educated in, owing to how the education system works (elaborated below).

The second metric is Medium of Instruction (MoI) in school. Most schools in India follow the Three Language Policy, wherein formal education is provided in English, Hindi, and the regional language (the idea and implementation of this policy, and its implications for regional minorities has been a matter of much public debate as well as scholarly work; for discussions see \alert{citation}). Sustained formal education in a language is a reliable predictor of LSRW (listening, speaking, reading, writing) skills. Since this study provides stimuli in the form of text, we wanted to ensure that all participants were sufficiently comfortable with reading and speaking both Bengali and English, to avoid confounds and uncontrolled variability due to reading/task effort.

With these considerations in mind, we only recruited those participants who had grown up in the state of West Bengal and lived there for the majority of their lives, had parents who spoke Bengali, and had received at least 5 years of formal education in Bengali (all respondents had received 5+ years of education in English and were studying university degrees that are taught in English), and self-reported proficiency in speaking, understanding, and reading Bengali.

A total of 28 bilingual speakers of Bengali and English (18 female) responded to the recruitment call. Using the inclusion criteria reported above, 10 volunteers (5 female, 5 male, age range 19 to 28) were invited to participate in a speech production study, and were compensated for their time. All participants reported normal hearing, and normal or corrected to normal vision. After the completion of the study, we followed up with a more detailed language background survey at a later date to learn more about the language profiles of the participants. This survey was a modified and consolidated version of three popular bilingualism profiling tools: the Bilingual Language Profile (BLP) \citep{blp}, the Language Experience and Proficiency Questionnaire (LEAP-Q) \citep{leap-q} which incorporates language history and self-assessed proficiency, and the Bilingual Switching Questionnaire (BSWQ) \citep{language_switching_questionnaire}. Together, these incorporate questions about language history, usage, attitudes, and switching experience.

In addition to Bengali and English, all participants reported knowing other languages, the most common being Hindi (n=10). Since this situation is representative of the present population and our study concerns only Bengali-English interaction, we only focused on these languages in the survey.
The average age of acquisition was 2.6 for English (range 0 i.e. since birth to 6) and 0.25 (range 0 to 2) for Bengali. Thus, all participants can be described as simultaneous bilinguals, having acquired both English and Bengali at a young age. They report being comfortable using both languages at a young age, with the average being 7.5 for English and 0.3 for Bengali (on a likert scale rating where 0=for as long as I can remember). As mentioned earlier, all participants had been in schools where Bengali and English were both used as Medium of Instruction (MoI). On average, participants had had more than 16 years (range 7 to 20+) of formal education in English and 11.8 years (range 8 to 15) in Bengali at the time of the study. As described in the inclusion criteria, all participants reported having spent over 15 years in a Bengali-speaking region. At the time of recording, they were living in or around the EFL University campus in Hyderabad, which has a linguistically diverse population. In terms of usage, participants show a range of language usage patterns. Almost 100\% of the participants report exclusively using Bengali with little to no English while interacting with family, and English at school/the workplace. However, language use with friends varies greatly, likely reflecting the language composition of the friend circle. 
The questionnaire asked participants to rate their proficiency in each language. All participants rated themselves as highly proficient in speaking and understanding both Bengali and English. While self-ratings for reading and writing English were at ceiling, ratings for Bengali literacy (reading and writing) showed some variation, with an average of 4.87 (on a 0-6 likert scale, range 3 to 6) for reading and 4.25 (range 3 to 6) for writing. This is a common situation for many educated sections of society, since institutions of higher education beyond the school level largely use English (for discussions about the language politics of higher education in India, see \alert{ \cite{ELE readings find}}). 
The language attitude section of the questionnaire reveals interesting information about the linguistic landscape of the population, and how people experience language in such a setting. When asked how much ``like themselves" participants feel while using each language, response were at ceiling for both languages. Similarly, most participants identified equally strongly with both an ``English-speaking" and a ``Bengali-speaking" culture, possibly because for most people, such linguistic-cultural identities are not clearly delineated and not exclusive of one another. 
These responses shed light on how multiple languages fit seamlessly into different facets of daily life to create a ``continuous" linguistic experience \alert{(c.f. plurality-- from Personal statement)}. In such a scenario, it is not surprising that language-switching is common, expected, as well as used is creative ways with communicative purpose. Responses to the section on language switching demonstrate this. On average, participants do not report being often unable to recall words in the target language while speaking Bengali (average rating 2.5 on a 1-6 likert scale where 1=never and 6=always) or English (average rating 2.3), indicating that switching is not strictly necessary for fulfilling lexical gaps. Indeed, participants do not report switching between languages without awareness (1.8) or control (1.25). In spite of this, participants report a proclivity to code-switch (rating 3.5). Importantly, a large number of participants report volition while switching languages (4.12), and many report it to be situationally motivated (3.12). These responses indicate that participants are highly aware of switching habits and move between languages for communicative purposes. This fact, combined with the linguistic situation described here and in section \ref{bengali_english_in_india}, are important to bear in mind while interpreting the results of a transfer study in the present population.

\subsection{Stimuli}
Twenty monosyllabic English words containing the vowels \nt{2} or \nt{\ae} were selected as target words. To minimize coarticulatory effects on the vowel, the pre-vocalic consonants were all voiced plosives (\nt{b} or \nt{d}). 10 monosyllabic English words that do not contain the target vowel served as filler items. We avoided English words that are already lexicalized loanwords in Bengali. %The list of words can be found in sec.\ref{appendix}.

\subsection{Procedure} \label{procedure}
Participants were recorded individually in a sound attenuated room. %, using a ... microphone type/description?
Productions were recorded in Praat \citep{boersma2016praat} at a 44.1K sampling rate. Each participant completed four iterations each of two tasks-- cued picture-naming and code-switching. Utterances in the two conditions (unilingual and mixed) were randomly interspersed within each iteration. In the unilingual condition, the target word was preceded by another English word in the utterance. In the mixed condition, the target was preceded by a Bengali word and thus produced as a switch from Bengali. Every target word appeared twice during a test block; once in a unilingual and once in a mixed condition. Participants alternated between the two tasks, and were allowed to take a short break after each test block. The tasks are described below:

@@

Carrier sentences were chosen such that the sound preceding the target word was uniform across all sentences (the final sound in \textit{yellow} is produced as [o] in Indian English), to minimize any possible differences due to coarticulatory effects on the target vowel. To ensure identical word order in both languages, we chose a specifier-noun construction, as this has a parallel structure in Bengali and English. This gave a uniform prosodic context for the target word across conditions. While presenting the mixed-language utterances, we used mixed orthography. \\

@@

All statistical analyses were performed in R \citep{r}. We used linear mixed effects models to test the effect of experimental conditions on vowel height (F1) and backness (F2) respectively. Since formants were extracted from five points over the course of articulating the vowel, time was coded numerically (1-5), and treated as an independent variable. This was scaled and centered prior to analysis and treated as a continuous predictor. We fitted models using the \emph{lme4} package \citep{lme4}, and used \emph{lmerTest} \citep{lmerTest} to obtain estimates of p values for individual variables. For all mixed models, the alpha criterion was set at $|t| > 2 $. The dependent variables were normalized F1 and F2, and we fitted separate models for these. The independent variables or fixed factors, along with their levels, are given in Table \ref{table variables}. Subject and Item were treated as random factors.\\

@@

Our study involved two IE vowels. Vowel category is expected to systematically affect formant frequencies (FFs). Thus, Vowel is a fixed effect in the models. The experimental treatment of interest is Context, i.e. whether the target item was presented in a unilingual (English) or mixed (Bengali) context. Our main research question (\bf{L2 vowels}) asks whether L2 vowel quality (formant frequencies) of proficient bilinguals is affected by the Context of utterance. We hypothesize that the direction of such shift in the mixed condition, if any, is towards a corresponding L1 category. Therefore, we expect the effect of Context on formant frequencies, i.e. the direction of shift, to be moderated by the vowel category (c.f. the discussion in Section \ref{asymmetry between sounds}, leading to our second research question \textbf{Asymmetry}). Thus, our main effect of interest is the interaction term Context*Vowel. This corresponds to the transfer pattern. Our third research question (\bf{Paradigm}) asks whether the language-mixing paradigm, or Task, affects the extent and direction of transfer. If this is the case, we expect Task to interact with the transfer term, giving a three way Task*Context*Vowel interaction. Since we measured FFs at five points in the vowel, Time is expected to affect the FF. However, the precise effect of formant dynamics is expected to depend on the vowel category. Thus, we expect the interaction of Vowel*Time to predict formant frequency. If transfer patterns change across time, then we should expect it to interact with the transfer term, giving a Context*Vowel*Time interaction.

@@

First, to test RQ 1 (\bf{L2 vowels}), we fitted a model with vowel*time, and vowel*context as fixed effects. The transfer term (vowel*context) is significant in this model (model coefficient/ unstandardized effect size ($\beta$)= -2.330e-01, standard error (SE)=  5.244e-02, t= -4.44, p=0.0002). Comparing this to a null model that lacks the transfer term shows that the full model is significantly better ($\chi^2(2) = 14.42, p=  0.0007$). To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction. The full model was better ($\chi^2(1) = 12.51, p= 0.0004$).

To test the hypothesis \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects. This three-way interaction term is significant ($\beta$= -1.955e-01, SE=  5.453e-02, t= -3.59, p= 0.0003). We also find a significant effect of task on FFs ($\beta$=3.300e-01, SE=2.668e-02, t=12.371, p=2e-16). Comparing this model to a null model that lacks the fixed effect Task ($\chi^2(4) = 572.61, p=  2.2e-16$) shows that the full model is better. Because Task has an independent effect on F1, and adding interactions increases model parameters, we fitted a second null model with Task as a separate fixed effect without the interaction term to test whether the interaction term is needed. However, the full model was still a better fit for the data ($\chi^2(3) = 23.82, p=  2.722e-05$). Next, we tested two other null models to verify that the full model with the three-way interaction term is indeed the optimal model for the data, by replacing the three-way interaction term with: (i) vowel*context + task*context (to test if task only interacts with context, rather than the transfer term vowel*context). The full model is better ($\chi^2(2) = 23.28, p= 8.78e-06$); (ii) vowel*context + task*vowel (to test if task only interacts with vowel, rather than the transfer term vowel*context). The full model is better ($\chi^2(2) = 13.36, p= 0.001$). This shows that the three-way interaction is the best fit for the data. Because an independent effect of task on FFs is unexpected (c.f. model interpretation below), we wanted to examine if the effect might be local, i.e. decrease over time. We added a task*time interaction term to the full model, which is significant ($\beta$= -1.216e-01, SE=  1.347e-02, t= -9.028, p= 2e-16 ) and improves model fit ($\chi^2(1) = 81.22, p= 2.2e-16$).

@@

Thus, the optimal model was: f1 $\sim$ time*vowel + vowel*context*task + time*task + (1 + context|word) + (1 + context + vowel|subject). Table \ref{table_f1} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model. \alert{p-vales smaller than 0.05 are in bold.}

\subsubsection*{Interpretation}
We extracted formant estimates from five uniformly spaced time points in the vowel: from the time at 5\% into the vowel (at the start of articulation) to 45\% (at the mid-part, or steady state, which is expected to correspond most closely to the articulatory target for the category). As expected, we see that F1 values are significantly predicted by Time in the model, reflecting the dynamic nature of articulation which causes acoustic variation over the course of time. The positive slope of the estimate for Time indicates that F1 increases, i.e. the two categories in the study show lowering, over the course of articulation. Moreover, the significant Time*Vowel interaction shows that the F1 trajectory is different for the two vowels. This is expected, as the categories are expected to present different articulatory targets.

Vowel is not a significant predictor of F1. This is initially surprising, given that the phonological category is expected to systematically affect vowel quality. However, a boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two IE categories \nt{2, \ae} and the extent of height difference between them. Note that these measures are only from the English (unilingual) condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching. Given that estimates for fixed effects are pooled across participants, this explains the lack of reliable differences between \nt{2} and \nt{\ae}.

The negative slope of the estimate for Context shows that F1 is lower, i.e. vowels are higher, in the unilingual (e) condition compared to the mixed (b) condition. Put more intuitively, vowels are lowered in the mixed condition. However, the lack significance indicates that using English words in mixed utterances does not lead to a uniform lowering/raising across vowels. Instead, the significant interaction between Vowel and Context signals that the extent and direction of shift is moderated by vowel category. There is an asymmetrical shift in the categories \nt{2} and \nt{\ae} in the mixed-language condition. Specifically, the Vowel*Context interaction has a negative slope, showing that the lowering is greater in \nt{2}. The lack of a significant Vowel*Context*Time interaction shows that transfer patterns do not change across time. This indicates a uniform scaling of formant trajectories in the mixed condition, suggesting that the underlying acoustic targets for these vowels are different during unilingual vs mixed utterances.

The results also reveal a significant effect of Task, such that the sentence task (code-switching) leads to higher F1 values in both vowels relative to the cued picture-naming task as: vowels are lowered in the code-switching task, independent of experimental condition. This is surprising, as we have no aproiri reason to expect a uniform difference in vowel quality across tasks. A significant interaction with Time with a positive slope of the estimate shows that this effect is local: it is strongest at the beginning of the vowel, and reduces over the course of the vowel. This is consistent with the interpretation that the unexpected Task effect is an artifact of the paradigm, possibly resulting from coarticulatory effects owing to the specific stimuli used in this study. Of greater interest to us is the significant Vowel*Context*Task interaction, which indicates a difference in transfer patterns across paradigms. Specifically, the negative slope of the estimate suggests that transfer in the F1 (height) dimension are greater in the code-switching (s) task compared to the picture-naming (p) task.\\

@@

Because an independent effect of task on FFs is unexpected (c.f. model interpretation below), we wanted to see if the effect might be local, i.e. decrease over time. We added a task*time interaction term to the full model, which is significant ($\beta$= 6.237e-02, SE=  8.876e-03, t= 7.027, p= 2.23e-12 ) and improves model fit ($\chi^2(1) = 49.271, p= 2.23e-12$).

@@
